---
title: "Metric and Model"
author: "Natalie Garrett"
date: "2023-12-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load in libraries
```{r libraries}
library(readr)
library(ggplot2)
library(dplyr)
library(tidyverse)
library(gganimate)
library(repr)
library(caTools)
library(MASS)
library(caret)
library(nortest)
```

get data
```{r getData}
games <- read_csv("games.csv")
players <- read_csv("players.csv")
plays <- read_csv("plays.csv")
tackles <- read_csv("tackles.csv")
tracking_week_1 <- read_csv("tracking_week_1.csv")
tracking_week_2 <- read_csv("tracking_week_2.csv")
tracking_week_3 <- read_csv("tracking_week_3.csv")
tracking_week_4 <- read_csv("tracking_week_4.csv")
tracking_week_5 <- read_csv("tracking_week_5.csv")
tracking_week_6 <- read_csv("tracking_week_6.csv")
tracking_week_7 <- read_csv("tracking_week_7.csv")
tracking_week_8 <- read_csv("tracking_week_8.csv")
tracking_week_9 <- read_csv("tracking_week_9.csv")
```

```{r function}
metric_creation <- function(tracking_data) {
# save plays with no penalty only and only needed columns
nonPenaltyPlays <- plays %>% filter(playNullifiedByPenalty != 'Y') %>% 
  dplyr::select(c(gameId, playId, defensiveTeam, ballCarrierId, 
           ballCarrierDisplayName, passResult))

# only want data that has tracking data of players and football
tracking_data <- tracking_data %>% group_by(gameId, playId, frameId) %>%
  filter(sum(displayName != 'football')>0) 

# data after tackle is irrelevant to scope
tracking_data <- tracking_data %>% group_by(gameId, playId) %>%
  mutate(last_frame = ifelse(frameId == max(frameId), "end", event)) %>% 
  ungroup()

tracking_data <- tracking_data %>% 
  fill(last_frame, .direction = 'up') %>% 
  filter(last_frame != 'end') %>%
  dplyr::select(-c(last_frame))

# join the dataframes and add metric of which side of ball player is on
tracking_data <- inner_join(tracking_data, nonPenaltyPlays, 
  by = c("gameId" = "gameId", "playId" = "playId")) %>%
  mutate(sideOfBall = ifelse(club == defensiveTeam, 'defense', 'offense'))

# standardize field and measurements as well as get metrics relevant to football and the horizontal/vertical movement of player
tracking_data$x <- ifelse(tracking_data$playDirection == "left",
                          120-tracking_data$x, tracking_data$x)
tracking_data$y <- ifelse(tracking_data$playDirection == "left", 
                          160/3 - tracking_data$y,tracking_data$y)
tracking_data$dir <- ifelse(tracking_data$playDirection == "left",
                            360-tracking_data$dir,tracking_data$dir)
tracking_data$o <- ifelse(tracking_data$playDirection == "left",
                          360-tracking_data$o, tracking_data$o)
tracking_data <- tracking_data %>% group_by(gameId, playId, frameId) %>%
  mutate(footballInPlay = sum(displayName == 'football')>0) %>%
  filter(footballInPlay) %>% 
  ungroup()

# make sure ball carrier tracking data is present
tracking_data <- tracking_data %>% group_by(gameId, playId, frameId) %>%
  filter(sum(displayName == ballCarrierDisplayName)==1) 

# add more metrics, including tracking x, y of ball carrier, separation between ball carrier, the x and y separation, vertical side of ball carrier (behind, in front, next to, and if player is chasing defender)
tracking_data$xs <- tracking_data$s*cos(tracking_data$dir*(pi/180))
tracking_data$ys <- tracking_data$s*sin(tracking_data$dir*(pi/180))
tracking_data$xa <- tracking_data$a*cos(tracking_data$dir*(pi/180))
tracking_data$ya <- tracking_data$a*sin(tracking_data$dir*(pi/180))
tracking_data$sideline_dist <- ifelse(tracking_data$y < 53.33/2,
                        tracking_data$y-0,53.33-tracking_data$y)

tracking_data <- tracking_data %>%
  group_by(gameId, playId, frameId) %>%
  mutate(xBallCarrier = x[displayName == ballCarrierDisplayName],
         yBallCarrier = y[displayName == ballCarrierDisplayName]) %>%
  ungroup() %>%
  group_by(gameId, playId) %>%
  mutate(separation = sqrt((x-xBallCarrier)^2+(y-yBallCarrier)^2),
         xSeparation = x-xBallCarrier, ySeparation = y-yBallCarrier) %>%
  ungroup() 

tracking_data$vSideOfPlayer <- ifelse(tracking_data$xSeparation > 0, 'behind',
                      ifelse(tracking_data$xSeparation < 0, 'front', 'nextTo'))
tracking_data$chasing = ifelse(tracking_data$vSideOfPlayer == 'behind', 1, 0)


# add more metrics, the speeds or accel in relation to ball carrier
tracking_data <- tracking_data %>% group_by(gameId, playId, frameId) %>%
  mutate(xBCrelsDef = xs - xs[displayName == ballCarrierDisplayName],
         yBCrelsDef = ys - ys[displayName == ballCarrierDisplayName],
         xBCrelaDef = xa - xa[displayName == ballCarrierDisplayName],
         yBCrelaDef = ya - ya[displayName == ballCarrierDisplayName])

tracking_data <- tracking_data %>%  group_by(gameId, playId, nflId) %>%
  mutate(deltaS = s- lag(s), deltaA = a- lag(a),
         deltaDir = dir- lag(dir), deltaO = o- lag(o),
         deltaXS = xs- lag(xs), deltaYS = ys- lag(ys),
         deltaXA = xa- lag(xa), deltaYA = ya- lag(ya))

# scramble - filter at run to tackle
# run - filter at handoff to tackle
# pass - filter pass outcome caught to tackle
passingPlay <- tracking_data %>%
  filter(passResult == 'C') %>%
  mutate(start_frame = ifelse(is.na(event), 0, 
                  ifelse(event == 'pass_outcome_caught', frameId, 0))) %>%
  group_by(gameId, playId, nflId) %>%
  mutate(irrelevant_frame = ifelse(frameId < max(start_frame), 
                                   TRUE, FALSE)) %>%
  ungroup() %>%
  filter(irrelevant_frame != TRUE) %>% 
  dplyr::select(-c(start_frame,irrelevant_frame)) %>%
  filter(separation < 5)

rushingPlay <- tracking_data %>%
  filter(is.na(passResult)) %>%
  mutate(start_frame = ifelse(is.na(event), 0, 
                  ifelse(event == 'handoff', frameId, 0))) %>%
  group_by(gameId, playId, nflId) %>%
  mutate(irrelevant_frame = ifelse(frameId < max(start_frame), 
                                   TRUE, FALSE)) %>%
  ungroup() %>%
  filter(irrelevant_frame != TRUE) %>% 
  dplyr::select(-c(start_frame,irrelevant_frame)) %>%
  filter(separation < 5)

scramblePlay <- tracking_data %>%
  filter(passResult == 'R') %>%
  mutate(start_frame = ifelse(is.na(event), 0, 
                  ifelse(event == 'run', frameId, 0))) %>%
  group_by(gameId, playId, nflId) %>%
  mutate(irrelevant_frame = ifelse(frameId < max(start_frame), 
                                   TRUE, FALSE)) %>%
  ungroup() %>%
  filter(irrelevant_frame != TRUE) %>% 
  dplyr::select(-c(start_frame,irrelevant_frame)) %>%
  filter(separation < 5)

tracking_data <- rbind(passingPlay, rushingPlay, scramblePlay)

return(tracking_data)
}

tracking_test_1 <- metric_creation(tracking_week_1)
tracking_test_2 <- metric_creation(tracking_week_2)
tracking_test_3 <- metric_creation(tracking_week_3)
tracking_test_4 <- metric_creation(tracking_week_4)
tracking_test_5 <- metric_creation(tracking_week_5)
tracking_test_6 <- metric_creation(tracking_week_6)
tracking_test_7 <- metric_creation(tracking_week_7)
tracking_test_8 <- metric_creation(tracking_week_8)
tracking_test_9 <- metric_creation(tracking_week_9)

tracking_data <- rbind(tracking_test_1, tracking_test_2,tracking_test_3,
                    tracking_test_4,
                    tracking_test_5, tracking_test_6,tracking_test_7,
                    tracking_test_8,tracking_test_9)

```

this model focus on plays where a completion was made and no penalty was called
prediction model only looks at when the players are within 5 yards of the ball carrier, considers only tackle data where a missed tackle, tackle or assist has occurred

```{r modelData}
only_tackles <- tackles %>%
  mutate(Tackle = ifelse((tackle == 1) | (assist == 1) ,
                         1, 0)) %>%
  filter(forcedFumble != 1)

defender_data <- tracking_data %>%
  filter(sideOfBall == 'defense') 

defender_data <- inner_join(defender_data, players, 
  by = c("nflId" = "nflId", "displayName" = "displayName")) %>%
  rename(tacklerPosition = position) %>%
  dplyr::select(-c(weight, height, birthDate, collegeName))


defender_data <- inner_join(defender_data, players, 
  by = c("ballCarrierId" = "nflId", 
         "ballCarrierDisplayName" = "displayName")) %>%
  rename(ballCarrierPosition = position) %>%
  dplyr::select(-c(weight, height, birthDate, collegeName))

defender_data$tacklerPosition <- as.factor(defender_data$tacklerPosition)
defender_data$ballCarrierPosition <-
  as.factor(defender_data$ballCarrierPosition)
defender_data$matchUpCombination <- paste(defender_data$tacklerPosition,
                              defender_data$ballCarrierPosition, sep = ', ')

defender_data <- left_join(defender_data, only_tackles, 
  by = c("gameId" = "gameId", "playId" = "playId", "nflId" = "nflId")) %>%
  distinct() %>%
  mutate(tackleMade = ifelse(is.na(Tackle), 0, Tackle)) %>%
  dplyr::select(-c(Tackle))

ballCarrier_data <- tracking_data %>%
  filter((displayName == ballCarrierDisplayName)) %>%
  dplyr::select(c(gameId, playId, frameId,s,a,dis,dir,o,xs,ys,xa,ya,
                 deltaS,deltaA,deltaXS,deltaYS,sideline_dist,deltaDir, 
                 deltaO)) %>%
  rename(BCs = s,BCa = a,BCdis = dis,BCdir=dir,BCo = o,
         BCxs=xs,BCys=ys,BCxa=xa,BCya=ya,BCdeltaS = deltaS,BCdeltaA = deltaA,
         BCdeltaXS=deltaXS,BCdeltaYS=deltaYS,BCsideline_dist=sideline_dist,
         BCdeltaDir=deltaDir, BCdeltaO=deltaO)

ballCarrier_data <- inner_join(ballCarrier_data, defender_data,
  by = c("gameId" = "gameId", "playId" = "playId", "frameId" = "frameId")) %>%
  dplyr::select(-c(s,a,dis,dir,o,xs,ys,xa,ya,
                 deltaS,deltaA,deltaXS,deltaYS,sideline_dist))
```


```{r normalityTesting}

sample_data <-defender_data %>%
  dplyr::select(c(s,a,dis,dir,o,xs,ys,xa,ya,xBCrelsDef,yBCrelsDef,
                 xBCrelaDef,yBCrelaDef,xSeparation,ySeparation,
                 chasing,deltaS,deltaA,deltaXS,deltaYS,sideline_dist,
                 deltaDir, deltaO))

for (i in colnames(sample_data)){
  if (ad.test(sample_data[[i]])[2] > 0.05) {
    print('defender')
    print(i)
    print(ad.test(sample_data[[i]])[2])
  }
}

sample_data <- ballCarrier_data %>%
  dplyr::select(c(BCs,BCa,BCdis,BCdir,BCo,BCxs,BCys,BCxa,BCya,
                 xBCrelsDef,yBCrelsDef,xBCrelaDef,yBCrelaDef,xSeparation,
                 ySeparation,chasing,BCdeltaS,BCdeltaA,BCdeltaXS,BCdeltaYS,
                 BCsideline_dist,deltaDir, deltaO))

for (i in colnames(sample_data)){
  if (ad.test(sample_data[[i]])[2] > 0.05) {
    print('BC')
    print(i)
    print(ad.test(sample_data[[i]])[2])
  }
}
```

```{r BCmodel}
set.seed(1)
split <- sample.split(ballCarrier_data$gameId, SplitRatio = 0.8)

train_reg <- subset(ballCarrier_data, split == "TRUE")
test_reg <- subset(ballCarrier_data, split == "FALSE")

QDA_mod <- qda(tackleMade ~ BCs+BCa+BCdis+BCdir+BCo+BCxs+BCys+BCxa+BCya+
                 xBCrelsDef+yBCrelsDef+BCdeltaDir+BCdeltaO+
                 xBCrelaDef+yBCrelaDef+xSeparation+ySeparation+
                 chasing+BCdeltaS+BCdeltaA+BCdeltaXS+BCdeltaYS+BCsideline_dist,
                    data = train_reg)

qda.pred1 = suppressWarnings(predict(QDA_mod,test_reg))

confusionMatrix(table(qda.pred1$class,test_reg$tackleMade))

log_mod <- glm(tackleMade ~ BCs+BCa+BCdis+BCdir+BCo+BCxs+BCys+BCxa+BCya+
                 xBCrelsDef+yBCrelsDef+BCdeltaDir+BCdeltaO+
                 xBCrelaDef+yBCrelaDef+xSeparation+ySeparation+
                 chasing+BCdeltaS+BCdeltaA+BCdeltaXS+BCdeltaYS+BCsideline_dist,
                    data = train_reg,
               family = "binomial")

summary(log_mod)
glm.probs1=predict(log_mod,test_reg,type="response")
n=dim(test_reg)[1]
glm.pred1=rep(0,n)
glm.pred1[glm.probs1>0.5]=1
confusionMatrix(table(glm.pred1,test_reg$tackleMade))

LDA_mod <- lda(tackleMade ~ BCs+BCa+BCdis+BCdir+BCo+BCxs+BCys+BCxa+BCya+
                 xBCrelsDef+yBCrelsDef+BCdeltaDir+BCdeltaO+
                 xBCrelaDef+yBCrelaDef+xSeparation+ySeparation+
                 chasing+BCdeltaS+BCdeltaA+BCdeltaXS+BCdeltaYS+BCsideline_dist,
                    data = train_reg)

lda.pred1 = suppressWarnings(predict(LDA_mod,test_reg))
confusionMatrix(table(lda.pred1$class,test_reg$tackleMade))

pred1 = suppressWarnings(round(predict(QDA_mod,
              ballCarrier_data)$posterior[,2], digits = 4)) 
ballCarrier_data$tackleProbability<- pred1

single_game <- ballCarrier_data %>% filter(gameId == 2022101605 
                                          & playId == 3778)
ggplot(single_game,
       aes(x=x,y=y, color = tackleProbability, shape = displayName )) + 
  geom_line() + facet_grid(playId ~ .)


ggplot(single_game,
       aes(x=frameId,y= tackleProbability, color = displayName )) + 
  geom_line() 

single_game <- ballCarrier_data %>% filter(playId == 1182	
                                          & gameId == 2022102000)

ggplot(single_game,
       aes(x=x,y=y, color = tackleProbability,
           shape = displayName )) + 
  geom_line() + facet_grid(playId ~ .)

ggplot(single_game,
       aes(x=frameId,y= tackleProbability, color = displayName )) + 
  geom_line() 

```


```{r Defmodel}
set.seed(1)
split <- sample.split(defender_data$gameId, SplitRatio = 0.8)

train_reg <- subset(defender_data, split == "TRUE")
test_reg <- subset(defender_data, split == "FALSE")

QDA_mod <- qda(tackleMade ~ s+a+dis+dir+o+xs+ys+xa+ya+xBCrelsDef+
                 yBCrelsDef+deltaDir+deltaO+
                 xBCrelaDef+yBCrelaDef+xSeparation+ySeparation+
                 chasing+deltaS+deltaA+deltaXS+deltaYS+sideline_dist,
                    data = train_reg)

qda.pred1 = suppressWarnings(predict(QDA_mod,test_reg))
confusionMatrix(table(qda.pred1$class,test_reg$tackleMade))

LDA_mod <- lda(tackleMade ~ s+a+dis+dir+o+xs+ys+xa+ya+xBCrelsDef+
                 yBCrelsDef+deltaDir+deltaO+
                 xBCrelaDef+yBCrelaDef+xSeparation+ySeparation+
                 chasing+deltaS+deltaA+deltaXS+deltaYS+sideline_dist,
                    data = train_reg)

lda.pred1 = suppressWarnings(predict(LDA_mod,test_reg))
confusionMatrix(table(lda.pred1$class,test_reg$tackleMade))

log_mod <- glm(tackleMade ~ s+a+dis+dir+o+xs+ys+xa+ya+xBCrelsDef+
                 yBCrelsDef+deltaDir+deltaO+
                 xBCrelaDef+yBCrelaDef+xSeparation+ySeparation+
                 chasing+deltaS+deltaA+deltaXS+deltaYS+sideline_dist,
                    data = train_reg,
               family = "binomial")

summary(log_mod)
glm.probs1=predict(log_mod,test_reg,type="response")
n=dim(test_reg)[1]
glm.pred1=rep(0,n)
glm.pred1[glm.probs1>0.5]=1
confusionMatrix(table(glm.pred1,test_reg$tackleMade))

pred1 = suppressWarnings(round(predict(QDA_mod, defender_data)$posterior[,2],
              digits = 4))
defender_data$tackleProbability <- pred1

single_game <- defender_data %>% filter(gameId == 2022101605 & playId == 3778)

ggplot(single_game,
       aes(x=x,y=y, group = displayName, 
           color = tackleProbability)) + #, shape = displayName 
  geom_point() + facet_grid(playId ~ .)


ggplot(single_game,
       aes(x=frameId,y= tackleProbability, color = displayName )) + 
  geom_line() 


single_game <- defender_data %>% filter(playId == 1182	
                                          & gameId == 2022102000)

ggplot(single_game,
       aes(x=x,y=y, group = displayName, 
           color = tackleProbability)) + 
  geom_point() + facet_grid(playId ~ .) #  , shape = displayName 

ggplot(single_game,
       aes(x=frameId,y= tackleProbability, color = displayName )) + 
  geom_line() 

```

perform joins differently and make probabilities own table to join later
```{r defineMistackle}

mistackle_data <- defender_data %>%
  filter(pff_missedTackle == 1)

mistackle_data_BC <- ballCarrier_data %>%
  filter(pff_missedTackle == 1) %>%
  rename(tackleProbabilityBC = tackleProbability) %>%
  dplyr::select(c(gameId, playId, nflId, displayName,
                  frameId, tackleProbabilityBC))

mistackle_data <- inner_join(mistackle_data, mistackle_data_BC, 
  by = c("gameId" = "gameId", "playId" = "playId", "nflId" = "nflId", 
         "frameId" = "frameId", "displayName" = "displayName"))

mistackle_data <- mistackle_data %>%
  group_by(gameId, playId, nflId) %>%
  mutate(mistackleTackle = ifelse(tackleProbability > 0.5 &
                        tackleProbabilityBC < 0.5, 1, 0)) %>%
  dplyr::select(c(gameId, playId, nflId, displayName,mistackleTackle)) %>%
  # , maxTackle,maxBCTackle
  distinct()

actual_mistackles <- tackles %>%
  filter(pff_missedTackle == 1 & gameId %in% mistackle_data$gameId) %>%
  dplyr::select(-c(gameId, playId, tackle, assist, forcedFumble)) %>%
  group_by(nflId) %>%
  mutate(tot_mistackles = sum(pff_missedTackle)) %>%
  dplyr::select(-c(pff_missedTackle)) %>%
  distinct()

mistackle_data <- mistackle_data %>%
  group_by(nflId)  %>%
  mutate(sumTackle = sum(mistackleTackle))  %>%
   dplyr::select(c(nflId, displayName, sumTackle)) %>%
  distinct()

mistackle_data <- inner_join(mistackle_data, actual_mistackles, 
                             by = c("nflId"="nflId")) %>%
  distinct() %>%
  mutate(MTOE = tot_mistackles - sumTackle) %>%
  rename(player = displayName, actual_missed_tackles = tot_mistackles) %>%
  dplyr::select(nflId, player, actual_missed_tackles, MTOE)
```


```{r countMistackles}
ggplot(mistackle_data, aes(x=MTOE)) + geom_bar() +
  labs(title = "Distribution of Missed Tackles Over Expected",
       x = "MTOE")

ggplot(mistackle_data, aes(x=MTOE, actual_missed_tackles)) + geom_point() +
  labs(title = "Distribution of Missed Tackles Over Expected",
       x = "MTOE", y = "Actual Number of Missed Tackles")
print(Sys.time() - start)
```
filter so only mistackle data, based on pff mistackle data

create more visuals
```{r MergeProb}
defender_probs <- defender_data %>% dplyr::select(c(gameId, playId,
  frameId, nflId, displayName, x, y, tackleProbability))

defender_probs$probs_type <- 'defender'

BC_probs <- ballCarrier_data %>% dplyr::select(c(gameId, playId,
  frameId, nflId, displayName, x, y, tackleProbability))

BC_probs$probs_type <- 'ball_carrier'

tackle_probs <- rbind(defender_probs, BC_probs)

single_game <- tackle_probs %>% filter(gameId == 2022101605 & nflId == 40017
                                          & playId == 3778)

single_game <- inner_join(single_game, plays,
                          by = c("gameId" = "gameId", "playId" = "playId"))

ggplot(single_game,
       aes(x=frameId,y= tackleProbability, color = displayName )) +
  geom_line() + facet_grid(probs_type ~ .) + labs(title = "Tackling probabilities during missed tackle by Tyrann Mathieu",caption = "(2:10) (Shotgun) J.Burrow pass short left to J.Chase for 60 yards, \n TOUCHDOWN. Pass 10, YAC 50")

single_game <- tackle_probs %>% filter(playId == 1182 & nflId == 44862 
                                          & gameId == 2022102000)


single_game <- inner_join(single_game, plays,
                          by = c("gameId" = "gameId", "playId" = "playId"))

ggplot(single_game,
       aes(x=frameId,y= tackleProbability, color = displayName )) +
  geom_line() + facet_grid(probs_type ~ .) + labs(title = "Tackling probabilities during missed tackle by Justin Evans",caption = "(9:56) (Shotgun) K.Murray pass short left to K.Ingram to ARZ 49 for \n 24 yards (C.Granderson).")

```

